---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
  knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# sspLNIRT

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

`sspLNIRT` is a sample size planning tool for item calibration with the Joint Hierarchical Model (JHM) of response accuracy and response time. It estimates the minimum sample size required to achieve a target accuracy (RMSE) of item parameter estimates under a specified data-generating process.

The package provides:

- **Precomputed results** for various design conditions, accessible instantly via a Shiny app and/or R package.
- **Custom Sample Size Estimation** via `optim_sample()` for design conditions outside the precomputed data.
- **Visualization functions** for inspecting parameter accuracy or bias, power curves, and implied response time and response accuracy distributions.

## Usage

The tool is available as an R package and as an interactive Shiny app.

### Web App

Use the app at [sebastian-lortz.shinyapps.io/sspLNIRT](https://sebastian-lortz.shinyapps.io/sspLNIRT/).

### Installation

You can install the latest version of the R package from GitHub:

```{r install-from-github, eval = FALSE, echo = TRUE}
# install devtools if needed
if (!requireNamespace("devtools")) {install.packages("devtools")}

# install from GitHub
devtools::install_github("sebastian-lortz/sspLNIRT")
```

### System Requirements

The `sspLNIRT` package was built under R version 4.5.2 using Apple clang version 16.0.0 (clang-1600.0.26.6) and GNU Fortran (GCC) 14.2.0. To compile R packages from source, install the appropriate toolchain:

- macOS: see <https://mac.r-project.org/tools/>
- Windows: see <https://cran.r-project.org/bin/windows/Rtools/>

### Run

Launch the Shiny app locally with:

```{r, eval = FALSE}
sspLNIRT::run_app()
```

## Documentation

Vignettes and full function documentation are available at [sebastian-lortz.github.io/sspLNIRT](https://sebastian-lortz.github.io/sspLNIRT/).

## Citation

Please cite `sspLNIRT` if you use it. To cite the software, use:

```{r echo = FALSE, results = "asis"}
print(citation("sspLNIRT"), style = "text")
```

Or copy the reference information to your BibTeX file:

```bibtex
@Manual{,
    title = {sspLNIRT: Sample Size Planning for Item Calibration using the Joint Hierarchical Model},
    author = {Sebastian A. J. Lortz},
    year = {2026},
    note = {R package version 0.0.0.9000},
    url = {https://github.com/sebastian-lortz/sspLNIRT},
  }
```

## Code of Conduct

I am open to feedback and new ideas.
Please mind the Contributor Code of Conduct.

## About

You are reading the doc about version: `r golem::pkg_version()`

This README has been compiled on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`.
